A Mechanised Library for Generic (Co)Programming

### 2024-03-07 paper/paper.tex
Recursive Anamorphisms

\subsection{Recursive Anamorphisms}
Anamorphisms specialise to an hylomorphism into an inductive data type.
Intuitively, this is because a recursive coalgebra can be applied only finitely
many times, so when this is applied to an anamorphism the only possible traces
it can produce are the finite ones.  In particular, we can show that
anamorphisms of type \coq{X -> GFix F} applied to a recursive \coq{F}-coalgebra
on \coq{X} are in one-to-one correspondence with hylomorphisms of
type \coq{X -> LFix F} on the same recursive \coq{F}-coalgebra and the
algebra \coq{In} for the inductive data type \coq{LFix F}.

To do this we first define a predicate on elements of coinductive data
structures which holds if and only if the element is finite:
\begin{coqcode}
Inductive FinF : GFix F -> Prop :=
| FinF_fold (x : GFix F) : (forall e, FinF (cont (g_out x) e)) -> FinF x.
\end{coqcode}
We can now define catamorphisms denoted by \coq{ccata_f_} from the
subtype \coq{{x : GFix F | FinF x}} of finitary elements of \coq{GFix F} to
any \coq{F}-algebra.
\begin{coqcode}
Definition ccata_f_ `{eA : setoid A} (g : Alg F A)
  : forall x : GFix F, FinF x -> A := fix f x H :=
    let hx := g_out x in
      g (MkCont (shape hx) (fun e => f (cont hx e) (FinF_inv H e))).
\end{coqcode}
In particular, we can take a catamorphism from the finitary elements
of \coq{GFix F} to the inductive data type \coq{LFix F} using
the \coq{F}-algebra \coq{l_in} and viceversa using catamorphism on the
restriction of \coq{g_in} to the finite elements of \coq{GFix}, which we denote
by \coq{lg_in}.
\begin{coqcode}
Lemma cata_ccata `{setoid A} (f : Alg F A) : cata f \o ccata l_in =e ccata f.
Lemma ccata_cata `{setoid A} (f : Alg F A) : ccata f \o cata lg_in =e cata f.
\end{coqcode}
Moreover, using this new initial algebra, we can compose catamorphisms and
recursive anamorphisms.

\begin{coqcode}
Corollary cata_ana_hylo `(F : Cont Sh P) `{setoid A} `{setoid B}
  (g : Alg F B) (h : RCoalg F A)
: cata g \o rana h =e hylo g h.
\end{coqcode}
This result will allow us to \emph{calculate} programs by providing small, easy
to verify implementations, and combining them into efficient recursive functions
by \emph{using Coq's proofs of hylomorphism laws as optimisation passes}.

This predicate is equivalent to the proposition that the anamorphism of such
coalgebras produce only finite trees:
\begin{coqcode}
Lemma ana_rec_term `{setoid A} (h : Coalg F A)
  : (forall x, FinF (ana h x)) <-> (forall x, RecF h x).
\end{coqcode}

\begin{coqcode}
Definition rana `{eA : setoid A} : RCoalg A ~> A ~> LFix F :=...

Lemma rana_univ A {eA : setoid A} (h : RCoalg A) (f : A ~> LFix F)
    : f =e rana h <-> f =e l_in \o fmap f \o h.
\end{coqcode}

\begin{coqcode}
Definition rana_f__ `{setoid A} (c : Coalg F A)
  : forall x : A, RecF c x -> LFix F
  := fix f (x : A) (H : RecF c x) :=
       let hx := c x in
       LFix_in (MkApp (shape hx) (fun e => f (cont hx e) (RecF_inv H e))).
\end{coqcode}

>>>>

\subsection{Recursive Anamorphisms}
Anamorphisms applied to recursive coalgebras specialise to hylomorphisms into an
inductive data type in the following way.  Intuitively, a recursive coalgebra
can be applied only finitely many times, therefore when this is applied to an
anamorphism the only possible traces it can produce from the seed function are
the finite ones. We denote this special kind of recursion scheme \emph{recursive
anamorphism}.  We can show that recursive anamorphisms of type \coq{X -> GFix F}
can also be given the type \coq{X -> LFix F}. Moreover, these anamorphisms are
exactly hylomorphisms on the recursive \coq{F}-coalgebra and the
algebra \coq{In} for the inductive data type \coq{LFix F}. This fact falls out
from the universality property of the hylomorphism and the fact that recursive
anamorphisms satisfy the same equation.

We now show how this is proven in the Coq development. First we define a
recursive anamorphism as a function into \coq{LFix F}:
\begin{coqcode}
Definition rana_f__ `{setoid A} (c : Coalg F A)
  : forall x : A, RecF c x -> LFix F
  := fix f (x : A) (H : RecF c x) :=
       let hx := c x in
       LFix_in (MkApp (shape hx) (fun e => f (cont hx e) (RecF_inv H e))).

Definition rana `{eA : setoid A} : RCoalg A ~> A ~> LFix F := ...
\end{coqcode}
We omit the boilerplate details to obtain the setoid morphism.
The recursive anamorphism has the following universality property:
\begin{coqcode}
Lemma rana_univ A {eA : setoid A} (h : RCoalg A) (f : A ~> LFix F)
    : f =e rana h <-> f =e l_in \o fmap f \o h.
\end{coqcode}
We can now show that every hylomorphism is composed by a recursive anamorphism
followed by a catamorphism:
\begin{coqcode}
Corollary cata_ana_hylo `(F : Cont Sh P) `{setoid A} `{setoid B}
  (g : Alg F B) (h : RCoalg F A)
: cata g \o rana h =e hylo g h.
\end{coqcode}
This result will allow us to \emph{calculate} programs by providing small, easy
to verify implementations, and combining them into efficient recursive functions
by using Coq's proofs of hylomorphism laws as optimisation passes.

\subsubsection{On the subtype of finite elements}
In this development we have defined recursive anamorphisms on inductive data
types.  We might have as well defined them on the subtype of finite elements of
coinductive data types using a predicate which states when an element of a
coinductive data type is finite:
\begin{coqcode}
Inductive FinF : GFix F -> Prop :=
| FinF_fold (x : GFix F) : (forall e, FinF (cont (g_out x) e)) -> FinF x.
\end{coqcode}
Now the subtype \coq{{x : GFix F | FinF x}} of finite elements for \coq{GFix F}
is isomorphic its corresponding the inductive data type \coq{LFix F}. This is
easy to see. We first define a catamorphism \coq{ccata_f_} from the
subtype \coq{{x : GFix F | FinF x}} of finitary elements of \coq{GFix F} to
any \coq{F}-algebra.
\begin{coqcode}
Definition ccata_f_ `{eA : setoid A} (g : Alg F A)
  : forall x : GFix F, FinF x -> A := fix f x H :=
    let hx := g_out x in
      g (MkCont (shape hx) (fun e => f (cont hx e) (FinF_inv H e))).
\end{coqcode}
We now prove this is isomorphic to the least fixed-point of the functor \coq{F}
We take the catamorphism from the finite elements of \coq{GFix F} to the
inductive data type \coq{LFix F} using the \coq{F}-algebra \coq{l_in}. Its inverse is the catamorphism on the restriction of \coq{g_in} to the finite elements of \coq{GFix}, which we denote by \coq{lg_in}.
The following lemmas prove the isomorphism:
\mpav{David, check this code works}
\begin{coqcode}
Lemma cata_ccata `{setoid A} : cata lg_in \o ccata l_in =e id.
Lemma ccata_cata `{setoid A} (f : Alg F A) : ccata l_in \o cata lg_in =e id.
\end{coqcode}
% This predicate is equivalent to the proposition that
% the anamorphism of such coalgebras produce only finite trees:
% \begin{coqcode}
% Lemma ana_rec_term `{setoid A} (h : Coalg F A)
%   : (forall x, FinF (ana h x)) <-> (forall x, RecF h x).
% \end{coqcode}
